---
title: "EMX105"
output:
  pdf_document: default
  html_document: default
date: "2024-05-11"
---

```{r}
# load functions
source("../R/tabll.R")
```

# 6,7
Test if there is any difference between the original value and the calculated one by trying out Exz.X as suggested. 
```{r test}
set.seed(121)
N <- 20
M <- 2
C <- 3
n <- 0.4*N
true.Z.vec <- as.vector(unlist(mapply(rep, c(1, 2), c(n, N-n))))
true.Z <- matrix(0, nrow = N, ncol = M)
true.eta <- round(prop.by.row(matrix(runif(M * C), nrow = M)), 2)
X <- matrix(nrow = N, ncol = C)
m.vec <- round(runif(N) * 100)
for(i in 1:N){
  true.Z[i, true.Z.vec[i]] <- 1
}
for(i in 1:N){
  X[i,] <- as.vector(rmultinom(1, size = m.vec[i], prob = true.Z[i,] %*% true.eta))
}

p.vec <- rep(1/M, M)
eta.mat <- prop.by.row(X)

for(it in 1:2){
expect.Z <- ExZ.X(p.vec, eta.mat, X)
expect.M <- Mparms(expect.Z, X)
p.vec <- expect.M$p
eta.mat <- expect.M$eta
}
# check ExZ.X
round(colMeans(expect.Z), 2)
# check eta.mat
round(expect.M$eta, 2)
true.eta
```
The Exz.X seems to return as expected.

# 11
```{r}
# read in data, and create a matrix X105
X105 <- read.csv("E:/FMPH291_SP24_C00_s3tu/X105.csv")
X105=as.matrix(X105, row.names = NULL)
```

```{r}
M <- 2
for(M in 2:10){
  X <- as.matrix(X105)
  N <- nrow(X)
  p.vec <- rep(1/M, M)
  eta.mat <- matrix(nrow = M, ncol = ncol(X))
  for(j in 1:M){
    eta.mat[j,] <- colMeans(prop.by.row(X[sample(1:N, size = 10),]))
    # print(eta.mat[j,])
  }

  # cat("M =", M)
  # cat("\n")
  ll.old <- tabll(p.vec, eta.mat, X)
  # cat("before update:", ll.old)
  # cat("\n")
  ll.new <- updatell(p.vec, eta.mat, X, 50)$ll
  # cat("after update:", ll.new)
  # cat("\n")
  # cat("\n")

  while(abs(ll.new - ll.old) > 0.001){
    cat("M =", M)
    cat("\n")
    cat("before update:", ll.old)
    cat("\n")
    cat("after update:", ll.new)
    cat("\n")
    cat("\n")
    ll.old <- ll.new
    ll.new <- updatell(p.vec, eta.mat, X, 50)$ll
  }
}
```
All updates of the loglikelihood finished within 50 iterations.  
It was not always increasing as M increased.
I think there are  probably 3 components as the log likelihood did not increase substantially after M=3.

\newpage
# Appendix: R Code for functions
```{r, echo=TRUE, eval=FALSE}
```

## function tabll
```{r tabll, echo=TRUE, eval=FALSE}
tabll <- function(p.vec, eta.mat, X){
  # X is a N by C matrix of counts
  # p.vec is a vector of length M
  N <- nrow(X)
  C <- ncol(X)
  M <- length(p.vec)
  if(ncol(eta.mat) != C) warning("input don't have compatible dimensions")
  if(nrow(eta.mat) != M) warning("input don't have compatible dimensions")
  m.vec <- rowSums(X)
  ll <- 1
  for(i in 1:N){
    lli <- 0
    for(j in 1:M){
      lli <- lli + p.vec[j] * dmultinom(X[i,], prob = eta.mat[j,])
      }
    ll <- ll * lli # observed data likelihood here
  }
  log(ll) # output log-likelihood
  }
```

## function ExZ.X
```{r ExZ.X, echo=TRUE, eval=FALSE}
# get proportion of row elements
prop.by.row <- function(X) sweep(X, 1, rowSums(X), "/")
ExZ.X <- function(p.vec, eta.mat, X){
  N <- nrow(X)
  C <- ncol(X)
  M <- length(p.vec)
  Ez <- matrix(nrow = N, ncol = M)
  for(i in 1:N){
    for(j in 1:M){
      Ez[i, j] <- p.vec[j] * dmultinom(X[i,], prob = eta.mat[j,])
    }
  }
  prop.by.row(Ez)
}
```

## function Mparms
```{r Mparms, echo=TRUE, eval=FALSE}
Mparms <- function(expect.Z, X){
  eta.mat <- prop.by.row(t(expect.Z) %*% X)
  p.vec <- colMeans(expect.Z)
  list(eta = eta.mat, p = p.vec)
}
```

## function updatell
```{r updatell, echo=TRUE, eval=FALSE}
updatell <- function(p.vec, eta.mat, X, n.iter = 10){
  # check if the input have compatible dimensions
  if(length(p.vec)!=dim(eta.mat)[1]|dim(eta.mat)[2]!=dim(X)[2])# M=M|C=C
  {warning("input don't have compatible dimensions")
    return(NULL)
  }
  for(it in 1:n.iter){
    expect.Z <- ExZ.X(p.vec, eta.mat, X)
    expect.M <- Mparms(expect.Z, X)
    p.vec <- expect.M$p
    eta.mat <- expect.M$eta
  }
  return(list(eta = eta.mat, p = p.vec, ll = tabll(p.vec, eta.mat, X)))
}
```

