---
title: "fmph291-hw2"
author: "Shengjia Tu"
date: "2024-04-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# The Pareto Distribution

An function dparetoR is produced to calculate the density for a Pareto distribution given $\alpha$, $\beta$ and x and saved it in “homeworks/R/dparetoR.R”. The code for this function is shown below.
```{r}
dparetoR <- function(x, a, b, logflag = FALSE){
  nx <- length(x)
  na <- length(a)
  nb <- length(b)
  n <- max(nx, na, nb)
  if (nx < n) x <- rep(x, length.out = n)
  if (na < n) a <- rep(a, length.out = n)
  if (nb < n) b <- rep(b, length.out = n)

  ld <- ifelse(a <= 0 | b <= 0, NaN, ifelse(x < a, -Inf, log(b) + b * log(a) - (b + 1) * log(x)))
  
  if(logflag){
    return(ld)
  }else{
    return(exp(ld))
  }
}
```

To avoid ambiguous, we introduced `logflag` instead of `log` as the parameter in the function.

# An Rcpp Approach 
An Rcpp function called dpareto that is vectorized by calling function paretodens repeatedly to construct a vector of results is produced. It is shown in the appendix and sourced below: 
```{r}
# load function
Rcpp::sourceCpp("../Rcpp/dpareto.cpp")
```

We first show some examples with the function paretodens which calculate the non-vectorized form of density for a Pareto distribution
```{r}
# examples for "paretodens"
paretodens(1, 1, 2)
paretodens(4, 1, 2)
# pareto <- Vectorize(paretodens)
```

The vectorized form function dpareto is tested with various cases: 
```{r}
# examples vectorized function, "dpareto"
set.seed(123)
x <- runif(5,0,10)
dpareto(x, 4.0, seq(3.0, 5.0, by= 0.75))
dparetoR(x, 4.0, seq(3.0, 5.0, by= 0.75))
dpareto(1.0, 2.0, NA)
dpareto(1000, 1000, 1000)
```

The rcpp dpareto code is compared to the R dparetoR code. They show the same result. Similarly, the two codes can be tested on other cases listed above. They show same results. 
```{r}
mat <- matrix(c(1, 1, 1, 1, 2, 2, 1, 3, 2, 2, 1, 2, 2, 2, 2, 2, 3, 3, 3, 1, 3, 3, 2, 3, 3, 3, 4, 4, 1, 4, 2, 3, 1, .5, .5, .5, -1, 1, 1, 2, -1, 0, 0, 0, 1), nrow = 3)
x <- mat[3, ]
alpha <- mat[1, ]
beta <- mat[2, ]
# paste0(beta*alpha^beta, "/", x^(beta+1))
# output <- strsplit("1, 1/4, 3/16, 1/2, 1, 24/81, 1/3, 2/3, 81/256, 1/4, 0, 1, NaN, NaN, NaN", ", ")[[1]]

# data.frame(x, alpha, beta, output)
x <- x[1:12]
alpha <- alpha[1:12]
beta <- beta[1:12]
all.equal(dparetoR(x, alpha, beta), dpareto(x, alpha, beta))
```

## Appendix
```{r, eval = FALSE, code = readLines("../Rcpp/dpareto.cpp"), engine = 'Rcpp'}
```