---
title: "Reject"
output:
  pdf_document: default
  html_document: default
date: "2024-05-20"
author: Shengjia Tu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- A brief explanation of how rejection sampling works: -->
<!-- The rejection sampling method generates sampling values from a target distribution $X$ with arbitrary probability density function $f(x)$ by using a proposal distribution $Y$ with probability density $g(x)$.  The idea is that one can generate a sample value from $X$ by instead sampling from $Y$ and accepting the sample $Y$ with probability $f(x)/(Mg(x))$, repeating the draws from $Y$ until a value is accepted. $M$ here is a constant and the finite bound on the likelihood ratio $f(x)/g(x)$. The  validation of this method is the envelope principle: when simulating the pair $(x,v=u*Mg(x))$, one produces a uniform simulation over the subgraph of $Mg(x)$. Accepting only pairs such that $u<f(x)/(Mg(x))$ then produces pairs $(x,y)$ uniformly distributed over the subgraph of  and thus, a simulation from $f(x)$. -->

<!-- In this homework,We know the density of proposal distribution(here it is Cauchy distribution) to sample from, and create an envelop which completely encloses the proposed distribution.  -->

<!-- The target function here is the dlnp function which returns the density of $\phi(x) exp(-exp(x))exp(xy)$.  -->

<!-- The envelop here is the random deviates function rlnp which returns samples from a target distribution whose density is proportional to $\phi(x) exp(-exp(x))exp(xy)$ (i.e. the R function dlnp) using a rejection sampler. In detail, the rlnp function should find a curve proportional to a Cauchy density (with scale as its scale parameter) that matches dlnp as nearly as possible and always is at least as high as dlnp. That curve is the envelope.  -->

## Rejection sampling
Let \(f\left( x\right) =ch\left( x\right) \) and \(g\left( x\right) \) denote two density functions with support \(\chi _{f}\) and \(\chi _{g}\), where $c$ is a normalizing constant for \(f\left( x\right)\), i.e., 

\[
c=\frac{1}{\int_{x\in \chi _{f}}h\left( x\right) dx}\quad \text{such that }%
\int_{x\in \chi _{f}}f\left( x\right) dx=c\int_{x\in \chi _{f}}h\left(
x\right) dx=1 
\]

We want to sample from \(f\left( x\right) \). \ However, it is not easy to sample from \(f\left( x\right) \), but it is easy to sample from \(g\left(x\right) \).

Assume that \(\chi _{f}\subseteq \chi _{g}\). \ Let 
\[
M=\sup_{x\in \chi _{f}}\frac{h\left( x\right) }{g\left( x\right) } 
\]
Then 
\[
h\left( x\right) \leq Mg\left( x\right) \quad \text{for all }x\in \chi _{f} 
\]

The rejection sampling is to sample from \( f\left( x\right) \) by sampling from \( g\left( x\right) \):

1. Simulate \(U\sim U\left( 0,1\right) \);

2. Simulate \(X\sim g\left( x\right) \);

3. If $X$ and $U$ satisfy: \
\[
U\leq \frac{h\left( X\right) }{Mg\left( X\right) }
\]
then accept $X$. \ Otherwise, discard $X$ and continue the above process until an accepted $X$. \

Let \(I\left( A\,\right) \) denote a set indicator. For the rejection sampling approach, the probability of accepting an $X$ is:
\begin{eqnarray*}
\Pr \left( X\text{ accepted}\right) &=&\Pr \left[ U\leq \frac{h\left(
X\right) }{Mg\left( X\right) }\right] \\
&=&E\left[ I\left( U\leq \frac{h\left( X\right) }{Mg\left( X\right) }%
\,\right) \right] \\
&=&E\left[ E\left\{ \left. I\left( U\leq \frac{h\left( X\right) }{Mg\left(
X\right) }\,\right) \right\vert X\right\} \right] \\
&=&E\left( \frac{h\left( X\right) }{Mg\left( X\right) }\right) \\
&=&\int_{x\in \chi _{g}}\frac{h\left( X\right) }{Mg\left( x\right) }g(x)dx \\
&=&\frac{1}{cM}\int_{x\in \chi _{g}}ch\left( X\right) dx \\
&=&\frac{1}{cM}\int_{x\in \chi _{f}}f\left( x\right) dx \\
&=&\frac{1}{cM}
\end{eqnarray*}

Thus the rejection sampling has effciency $\frac{1}{cM}$. \ 

To show that the accepted $X$ follows the distribution \(f\left( x\right) \),
we have: 
\begin{eqnarray*}
\Pr (X &\leq &t\mid X\text{ accepted})=\frac{\Pr \left( X\leq t,X\text{
accepted}\right) }{\Pr \left( X\text{ accepted}\right) } \\
&=&\frac{\Pr \left( X\leq t,X\text{ accepted}\right) }{\frac{1}{cM}} \\
&=&cME\left[ I\left( X\leq t\right) I\left( U\leq \frac{h(X)}{Mg(X)}\right) %
\right]  \\
&=&cME\left\{ E\left[ \left. I\left( X\leq t\right) I\left( U\leq \frac{h(X)%
}{Mg(X)}\right) \right\vert \,X\right] \right\}  \\
&=&cME\left\{ I\left( X\leq t\right) E\left[ \left. I\left( U\leq \frac{h(X)%
}{Mg(X)}\right) \right\vert \,X\right] \right\}  \\
&=&cME\left[ I\left( X\leq t\right) \frac{h(X)}{Mg(X)}\right]  \\
&=&M\int_{x\in \chi _{g}}I\left( x\leq t\right) \frac{ch\left( x\right) }{%
Mg\left( x\right) }g(x)dx \\
&=&\int_{x\in \chi _{g}}I\left( x\leq t\right) ch\left( x\right) dx \\
&=&\int_{x\in \chi _{g}}I\left( x\leq t\right) f\left( x\right) dx \\
&=&F\left( t\right) 
\end{eqnarray*}

For Homework 8, we want to sample from the density function:\ 
\[
f(x;\mu ,\sigma ,y)=c\times \phi (x;\mu ,\sigma )\times \exp (-\exp
(x))\times \frac{\exp (x)^{y}}{y!}
\]
where $c$ is a normalizing constant and $\phi (x;\mu ,\sigma )$ denotes the
density of the normal \(N\left( \mu ,\sigma \right) \). \ The
parameters of the above distribution are \(\mu ,\sigma \) and \(y\). \ 

We want to sample $f(x;\mu ,\sigma ,y)$ through sampling a Cauchy density:\ 
\[
g\left( x;x_{0},\lambda \right) =\left( \pi \lambda \left( 1+\frac{\left(
x-x_{0}\right) ^{2}}{\lambda ^{2}}\right) \right) ^{-1}
\]
where \(x_{0}\) is the location and \(\lambda \) is the scale parameter. \ 

For example, given $\mu =0,\sigma =1$ and $y=2$, the density $f(x;0,1,2)$
becomes:
\[
f(x;0,1,2)=c\times \phi (x;0,1)\times \exp (-\exp (x))\times \frac{\exp
(x)^{2}}{2!}
\]

To use the Cauchy density as an envelop function, we can set $\lambda =1$
and find $M$ such that for all $x\in R$: 
\[
M\times g\left( x;x_{0},1\right) \geq f(x;0,1,2)/c=\phi (x;0,1)\times \exp
(-\exp (x))\times \frac{\exp (x)^{2}}{2!}
\]

To find $M$, we first determine the mode of $f(x;0,1,2)/c$. \ If $x_{m}$ is
the mode of $f(x;0,1,2)/c$, we then set $x_{0}=x_{m}$ in the Cauchy density \(g\left( x;x_{0},1\right) \). \ We can then determine $M$ by:      
\[
M=\frac{\phi (x_{m};0,1)\times \exp (-\exp (x_{m}))\times \frac{\exp
(x_{m})^{2}}{2!}}{g\left( x;x_{m},1\right) }
\]

The plots of $\phi (x;0,1)\times \exp (-\exp (x))\times \frac{\exp (x)^{2}}{2!}$ and \(Mg\left( x;x_{m},1\right) \) are roughly `bell-shaped' curves with
the $\phi (x;0,1)\times \exp (-\exp (x))\times \frac{\exp (x)^{2}}{2!}$
curve hardly ever (or never) exceeding the \(Mg\left( x;x_{m},1\right) \)
curve, but very close to it at the mode. \

When sampling \(\left( X,U\right) \) uniformly from the area under the envelop
curve \(Mg\left( x;x_{m},1\right) \), points \(\left( X,U\right) \) that lie
below the curve of $\phi (x;0,1)\times \exp (-\exp (x))\times \frac{\exp
(x)^{2}}{2!}$ satisfy:
\[
U\leq \frac{\phi (X;0,1)\times \exp (-\exp (X))\times \frac{\exp (X)^{2}}{2!}%
}{Mg\left( X;x_{m},1\right) }
\]

Similarly, points \(\left( X,U\right) \) that satisfy the above will lie below
the curve of $\phi (x;0,1)\times \exp (-\exp (x))\times \frac{\exp (x)^{2}}{2!}$. \ 

## log-Normal Poisson distribution

The density of LNP distribution:  
$$
f(x;y)=\phi(x)\exp(-\exp(x))\exp(xy)
$$

```{r}
dlnp <-
function(x,y ,log = FALSE) {
val <-
dnorm( x, log = TRUE ) +
dpois( y, exp( x ), log = TRUE)
if (log) val else exp( val )
}
```


Using codes below in console(including commented lines), the mode of the density can be found roughly with cursor.
```{r}
# windows()
curve(dlnp(x,2),-5,5)
grid()
# locator(1)
# $x
# 0.4269226
# $y
# 0.09201661
```

Similarly, we find the mode coordinates for the Cauthy distribution roughly using the cursor.
```{r, eval=FALSE}
# windows()
curve(dcauchy(x,0.4269226,1),-5,5)
grid()
# locator(1)
# $x
# 0.3949434
# $y
# 0.3180528
```

### Envelope
Envelope is a scaled Cauchy density that approximates a LNP density by timing the relative height of a LNP density curve to a Cauchy density curve.  

We rescale the height of the Cauthy density by the ratio of the height of dlnp to the height of the Cauthy.

```{r}
curve(dlnp(x,2),-5,5)
curve(0.09201661/0.3180528*dcauchy(x,0.4269226,1),-5,5, add = T, col="red")
```
From the plot above, two roughly ‘bell-shaped’ curves with the dlnp curve hardly ever (or never) exceeding the other curve, but very close to it at the mode.


## rlnp & rlnpPlot
The code of function rlnp() & rlnpPlot() is attached in the appendix.
```{r,echo=FALSE}
rlnp <- function(n, y, scale = 1){
  # a <- -1 + log( max(1, y) )
  # b <- log( max(1, y) )
  # x0 <- optimize(function(x){dlnp(x, y)}, c(a, b), maximum = T)$maximum
  # M <- exp(optimize(function(x){dlnp(x0,y,log=T)-dcauchy(x, x0, scale,log=T)}, x0+c(-2, sqrt( .Machine$double.eps )), maximum = T)$objective+1e-06)
  x0 <- optimize(function(x){dlnp(x, y)}, c(-5, 5), maximum = T)$maximum
M <- dlnp(x0,y)/optimize(function(x){dcauchy(x, x0, scale)}, c(-5, 5), maximum = T)$objective
  x <- c()
  yunif <- c()
  dratio <- c()
  keep <- c()
  n_ <- 10*n
  while(sum(keep) < n){
    x_ <- rcauchy(n_)
    yunif_ <- runif(n_)
    dratio_ <- dlnp(x_, y) / (M*dcauchy(x_, x0, scale))
    keep_ <- yunif_ <= dratio_
    x <- c(x, x_)
    yunif <- c(yunif, yunif_)
    dratio <- c(dratio, dratio_)
    keep <- c(keep, keep_)
  }
  i.end <- max(which(keep)[1:n])
  return(list(x = x[1:i.end], yunif = yunif[1:i.end], dratio = dratio[1:i.end], keep = keep[1:i.end], x0 = x0, M = M, y = y, scale = scale))
}

rlnpPlot <- function(obj){
  with(obj, {
    curve( M * dcauchy( x , x0, scale), min(x[ keep ]), max(x[ keep ], lty = 4), ylab="", xlab = "")
    curve( dlnp( x , y), min(x[ keep ]), max(x[ keep ]), col = 2, add = TRUE, ylab="", xlab = "")
    points( x[ keep ], yunif[ keep ] * (dcauchy( x[ keep ] , x0, scale) * M), pch=".", col=2, ylab="", xlab = "")
    points( x[ ! keep ], yunif[ ! keep ] * (dcauchy( x[ ! keep ] , x0, scale) * M), pch=".", ylab="", xlab = "")
    legend( "topright", col = 1:2, lty = rep( 1, 2 ), legend = c( "denvelope/rejected", "dlnp/accepted"))
    title(main = paste0("dlnp(x, ", y, ")"), sub = paste0("denvelope(x) = ", round(M,2), " * dcauchy(x,", round(x0,2), ",", round(scale,2), ")", "\n", "fraction of samples rejected = ", round(1-sum(keep)/length(keep),2)))
  })
}
```

When the y value is 0, the envelope does not 100% cover the LNP, since there were more than 5700 dratio values exceeding 1(see below). So the sampler
has some flaw.
```{r}
lst.lnp <- rlnp(10000, 0)
system.time({
  lst.lnp <- rlnp(10000, 0)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r, echo=FALSE}
dratio_ex1 <- function(obj, nmax=20){
  dratio_ex <- obj$dratio[which(lst.lnp$dratio>1)]
if(length(dratio_ex) == 0){
  cat("No dratio exceeds 1.0")
}else if(length(dratio_ex) > nmax){
  cat(length(dratio_ex), "dratio values exceeds 1.0")
}else{
  cat(length(dratio_ex), "dratio values exceeds 1.0:", paste(dratio_ex, sep = ", "))
}
}
```

```{r}
dratio_ex1(lst.lnp)
```


When the y value is 3, the envelop covers dlnp 100% percent.
The higher(compared to the previous case with y=0) rejection rate is due to the scaling factor CM.
```{r}
lst.lnp <- rlnp(10000, 3)
system.time({
  lst.lnp <- rlnp(10000, 3)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```

<!-- include enough plots to explain what rlnp(10000,0) and rlnp(10000,3) show. -->
When the y value is 10, the envelop covers the dlnp 100%. The higher rejection rate(compared to the case with y=0,3) is again due to the scaling  factor cM.

```{r}
lst.lnp <- rlnp(10000, 10)
system.time({
  lst.lnp <- rlnp(10000, 10)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```

When the y = 100, the envelop covers dlnp 100%, the rejection rate in this case is extremely high, almost 1.

```{r}
lst.lnp <- rlnp(10000, 100)
system.time({
  lst.lnp <- rlnp(10000, 100)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```

If the scale parameter change to 2, the rejection
rate is also close to 1. By Comparing the plots, more samples seems to be rejected when the scale is 2.

```{r}
lst.lnp <- rlnp(10000, 100, 2)
system.time({
  lst.lnp <- rlnp(10000, 100, 2)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```


When the scale changed to 0.1, the envelop does not completely cover dnlp.

```{r}
lst.lnp <- rlnp(10000, 100, 0.1)
system.time({
  lst.lnp <- rlnp(10000, 100, 0.1)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```


When the scale change to 0.2, the envelop covers 100% of dlnp, but the rejection rate is close to 1. The close to 1 rejection rate is again due to the low sampling efficiency 1/cM

```{r}
lst.lnp <- rlnp(10000, 100, 0.2)
system.time({
  lst.lnp <- rlnp(10000, 100, 0.2)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```


When the scale changed to 0.15, the rejection rate is a bit lower. And we choose scale = 0.15 as
the best scale for y=100.

```{r}
lst.lnp <- rlnp(10000, 100, 0.15)
system.time({
  lst.lnp <- rlnp(10000, 100, 0.15)
})
rlnpPlot(lst.lnp)
system.time(rlnpPlot(lst.lnp))
```

```{r}
dratio_ex1(lst.lnp)
```

The beginning of this report (pp.1-2) provides a brief explanation of why and how rejection sampling works. 
The target function here is LNP and the envelop function is the Cauthy.

<!-- Check the values of the dratio element for each run of rlnp. If any exceeds 1.0 report that fact and the -->
<!-- value observed. Offer a judgment as to whether a serious flaw in your sampler has been revealed and explain why you made that judgment. -->
```{r}

```

\newpage

## Source codes of functions
### dlnp
```{r, eval=FALSE}
dlnp <-
function(x,y ,log = FALSE) {
val <-
dnorm( x, log = TRUE ) +
dpois( y, exp( x ), log = TRUE)
if (log) val else exp( val )
}
```

### rlnp 
```{r}
rlnp <- function(n, y, scale = 1){
  x0 <- optimize(function(x){dlnp(x, y)}, c(-5, 5), maximum = T)$maximum
M <- dlnp(x0,y)/optimize(function(x){dcauchy(x, x0, scale)}, c(-5, 5), maximum = T)$objective
  x <- c()
  yunif <- c()
  dratio <- c()
  keep <- c()
  n_ <- 10*n
  while(sum(keep) < n){
    x_ <- rcauchy(n_)
    yunif_ <- runif(n_)
    dratio_ <- dlnp(x_, y) / (M*dcauchy(x_, x0, scale))
    keep_ <- yunif_ <= dratio_
    x <- c(x, x_)
    yunif <- c(yunif, yunif_)
    dratio <- c(dratio, dratio_)
    keep <- c(keep, keep_)
  }
  i.end <- max(which(keep)[1:n])
  return(list(x = x[1:i.end], yunif = yunif[1:i.end], dratio = dratio[1:i.end], keep = keep[1:i.end], x0 = x0, M = M, y = y, scale = scale))
}
```

### rlnpPlot
```{r,eval=FALSE}
rlnpPlot <- function(obj){
  with(obj, {
    curve( M * dcauchy( x , x0, scale), min(x[ keep ]), max(x[ keep ], lty = 4), ylab="", xlab = "")
    curve( dlnp( x , y), min(x[ keep ]), max(x[ keep ]), col = 2, add = TRUE, ylab="", xlab = "")
    points( x[ keep ], yunif[ keep ] * (dcauchy( x[ keep ] , x0, scale) * M), pch=".", col=2, ylab="", xlab = "")
    points( x[ ! keep ], yunif[ ! keep ] * (dcauchy( x[ ! keep ] , x0, scale) * M), pch=".", ylab="", xlab = "")
    legend( "topright", col = 1:2, lty = rep( 1, 2 ), legend = c( "denvelope/rejected", "dlnp/accepted"))
    title(main = paste0("dlnp(x, ", y, ")"), sub = paste0("denvelope(x) = ", round(M,2), " * dcauchy(x,", round(x0,2), ",", round(scale,2), ")", "\n", "fraction of samples rejected = ", round(1-sum(keep)/length(keep),2)))
  })
}
```

### dratio_ex1--reporting number of dratio>1 and their values
```{r, eval=FALSE}
dratio_ex1 <- function(obj, nmax=20){
  dratio_ex <- obj$dratio[which(lst.lnp$dratio>1)]
if(length(dratio_ex) == 0){
  cat("No dratio exceeds 1.0")
}else if(length(dratio_ex) > nmax){
  cat(length(dratio_ex), "dratio values exceeds 1.0")
}else{
  cat(length(dratio_ex), "dratio values exceeds 1.0:", paste(dratio_ex, sep = ", "))
}
}
```