---
title: "trimean"
author: Shengjia Tu
date: "5/31/2024"
output:
  pdf_document:
    latex_engine: xelatex
---

## R Markdown

The trimean is an estimate of location based on the median and the quartiles
that is constructed as a weighted average with the median weighted twice as
heavily as each quartile. \ For Homework 9, we use simulation study to
examine bias and reduction of variance by the control-variate method. \


Let $X_{i}$ denote an i.i.d. sample from a given distribution \(f\left(
x\right) \) \(\left( 1\leq i\leq 9\right) \) and let $X_{\left( i\right) }$
denote the order statistic, i.e.,
\[
X_{\left( 1\right) }\leq X_{\left( 2\right) }\leq \cdots \leq X_{\left(
9\right) }
\]
The trimean is defined as:
\[
\widehat{\theta }=\frac{X_{\left( 3\right) }+2X_{\left( 5\right) }+X_{\left(
7\right) }}{4}
\]

The naive simulation estimate is to estimate the mean of $X_{i}$ using the
trimean. \

To implement a control-variate approach, we need to find a control variate W and compute the following estimate:\
\[
\widetilde{\theta }=\widehat{\theta }-b\left( W-E\left( W\right) \right)
\]
where $b$ is a known constant. \ By taking expectation on both sides, we
have:
\begin{eqnarray*}
E\left( \widetilde{\theta }\right) &=&E\left( \widehat{\theta }\right) -E%
\left[ b\left( W-E\left( W\right) \right) \right] \\
&=&\delta -bE\left( W-E\left( W\right) \right) \\
&=&\delta
\end{eqnarray*}
Thus the control-variate estimator $\widetilde{\theta }$ has the same mean
as the trimean. \ Note that if the trimean is biased, $\delta \neq 0$. \
For Homework 9, we choose the sample mean \(\overline{X}=\frac{1}{9}
\sum_{i=1}^{9}X_{i}\) as the control variate \(W\). \ This control variate $%
\overline{X}$ is highly correlated with the trimean $\widehat{\theta }$. \
Thus we choose $b=1$.\ \ Since $E\left( \overline{X}\right) =0$ for all
three distributions for the simulation study, the\ trimean with this
control-variate is given by:\
\[
\widetilde{\theta }=\widehat{\theta }-\overline{X}
\]

We use Monte Carlo simulation to compare the two estimators \(\widetilde{\theta }\) and \(\widehat{\theta }\) in terms of bias, \(E\left( \widetilde{\theta }\right) \) and \(E\left( \widehat{\theta }\right) \), and variance \(Var\left( \widetilde{\theta }\right) \) and \(Var\left( \widehat{\theta }\right) \). \

Let $N$ denote the Monte Carol sample size. \ Let $\widetilde{\theta }%
^{\left( k\right) }$ and $\widehat{\theta }^{\left( k\right) }$ denote the
two trimean estimators from the $k$th Monte Carlo sample \(\left( 1\leq k\leq
N\right) \). \ Then we estimate the bias and variance of each estimator by:\
\begin{eqnarray*}
\text{bias} &:&\widehat{E}\left( \widetilde{\theta }\right) =\frac{1}{N}%
\sum_{i=1}^{N}\widetilde{\theta }^{\left( k\right) },\quad \widehat{E}\left(
\widehat{\theta }\right) =\frac{1}{N}\sum_{i=1}^{N}\widehat{\theta }^{\left(
k\right) } \\
\text{variance} &:&\widehat{Var}\left( \widetilde{\theta }\right) =\frac{1}{%
N-1}\sum_{i=1}^{N}\left( \widetilde{\theta }^{\left( k\right) }-\widehat{E}%
\left( \widetilde{\theta }\right) \right) ^{2},\quad \widehat{Var}\left(
\widetilde{\theta }\right) =\frac{1}{N-1}\sum_{i=1}^{N}\left( \widehat{%
\theta }^{\left( k\right) }-\widehat{E}\left( \widehat{\theta }\right)
\right) ^{2}
\end{eqnarray*}


Construct samples from Standard normal distribution as below: 

```{r}
#sample size is 9
n <- 9
# simulation 10000 times
N <- 10000

# mean of standard normal distribution 
mu_ <- 0
# standard deviation of standard normal distribution 
sd_ <- 1

set.seed(321)
#construct the trimean function that is an estimate of location based on the median and the quartiles that is constructed as a weighted average with the median weighted twice as heavily as each quartile.
Xtrimean.snorm <- rep(NA, N)# naive, without control variate
Xtrimean2.snorm <- rep(NA, N) # with control variate
for(j in 1:N){
  Xn <- rnorm(n, mu_, sd_)
  Xn <- sort(Xn)
  Xtrimean.snorm[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4
  Xtrimean2.snorm[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4-mean(Xn)# with control covariate
}
# bias of naive trimean
E.Xtrimean.snorm <- mean(Xtrimean.snorm)
# bias of trimean with control variate
E.Xtrimean2.snorm <- mean(Xtrimean2.snorm)
# SE of naive trimean
Var.Xtrimean.snorm <- var(Xtrimean.snorm)
# (Var.Xtrimean2 <- var(Xtrimean2))
# standard errors of estimates
SE.Var.Xtrimean.snorm <- sqrt(sum((Xtrimean.snorm-E.Xtrimean.snorm)^2)/(N-1))
# SE of trimean with control variate
Var.Xtrimean2.snorm <- var(Xtrimean2.snorm)
SE.Var.Xtrimean2.snorm <- sqrt(Var.Xtrimean2.snorm)
```

Then we construct t-distribution with 4 degrees of freedom: 

```{r}
#sample size is 9
n <- 9
# simulation 10000 times
N <- 10000

# mean of standard normal distribution 
mu_ <- 0
# standard deviation of standard normal distribution 
sd_ <- 1

set.seed(321)
#construct the trimean function that is an estimate of location based on the median and the quartiles that is constructed as a weighted average with the median weighted twice as heavily as each quartile.
Xtrimean.t4 <- rep(NA, N)# naive, without control variate
Xtrimean2.t4 <- rep(NA, N) # with control variate
for(j in 1:N){
  Xn <- rt(n, df=4)
  Xn <- sort(Xn)
  Xtrimean.t4[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4
  Xtrimean2.t4[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4-mean(Xn)# with control covariate
}
# bias of naive trimean
E.Xtrimean.t4 <- mean(Xtrimean.t4)
# bias of trimean with control variate
E.Xtrimean2.t4 <- mean(Xtrimean2.t4)
# SE of naive trimean
Var.Xtrimean.t4 <- var(Xtrimean.t4)
SE.Var.Xtrimean.t4 <- sqrt(Var.Xtrimean.t4)
# SE of trimean with control variate
Var.Xtrimean2.t4 <- var(Xtrimean2.t4)
SE.Var.Xtrimean2.t4 <- sqrt(Var.Xtrimean2.t4)
```

Next we construct a t-distribution with 10 degrees of freedom: 

```{r}
#sample size is 9
n <- 9
# simulation 10000 times
N <- 10000

# mean of standard normal distribution 
mu_ <- 0
# standard deviation of standard normal distribution 
sd_ <- 1

set.seed(321)
#construct the trimean function that is an estimate of location based on the median and the quartiles that is constructed as a weighted average with the median weighted twice as heavily as each quartile.
Xtrimean.t10 <- rep(NA, N)# naive, without control variate
Xtrimean2.t10 <- rep(NA, N) # with control variate
for(j in 1:N){
  Xn <- rt(n, df=10)
  Xn <- sort(Xn)
  Xtrimean.t10[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4
  Xtrimean2.t10[j] <- (Xn[3]+2*Xn[5]+Xn[7])/4-mean(Xn)# with control covariate
}
# bias of naive trimean
E.Xtrimean.t10 <- mean(Xtrimean.t10)
# bias of trimean with control variate
E.Xtrimean2.t10 <- mean(Xtrimean2.t10)
# SE of naive trimean
Var.Xtrimean.t10 <- var(Xtrimean.t10)
SE.Var.Xtrimean.t10 <- sqrt(Var.Xtrimean.t10)
# SE of trimean with control variate
Var.Xtrimean2.t10 <- var(Xtrimean2.t10)
SE.Var.Xtrimean2.t10 <- sqrt(Var.Xtrimean2.t10)
```

## Question 1
```{r}
cbind(E.Xtrimean.snorm, E.Xtrimean.t4, E.Xtrimean.t10)
table1=matrix(c(Var.Xtrimean.snorm, SE.Var.Xtrimean.snorm,
              Var.Xtrimean.t4, SE.Var.Xtrimean.t4,
              Var.Xtrimean.t10, SE.Var.Xtrimean.t10),
              nrow=3, ncol=2, byrow = TRUE)
colnames(table1)=c("Variance","SE")
rownames(table1)=c("snorm","t4","t10")
table1
```
The results for the bias show that the trimean estimator is unbiased, since the biases  are very small for all three distributions. The results for the variance/standard errors show that the trimean estimator has slightly smaller variance for the standard normal than the two t-distributions. Also, the variance for the t-distribution with higher degrees of freedom is closer to  the standard normal.


## Question 2
```{r}
percentage.snorm=100*(Var.Xtrimean.snorm-Var.Xtrimean2.snorm)/Var.Xtrimean.snorm
percentage.t4=100*(Var.Xtrimean.t4-Var.Xtrimean2.t4)/Var.Xtrimean.t4
percentage.t10=100*(Var.Xtrimean.t10-Var.Xtrimean2.t10)/Var.Xtrimean.t10
table2=matrix(c(Var.Xtrimean.snorm, Var.Xtrimean2.snorm, percentage.snorm,
              Var.Xtrimean.t4, Var.Xtrimean2.t4, percentage.t4,
              Var.Xtrimean.t10, Var.Xtrimean2.t10, percentage.t10),
              nrow=3, ncol=3, byrow = TRUE)
colnames(table2)=c("Variance","Variance with control","percent of reduction")
rownames(table2)=c("snorm","t4","t10")
table2
```
Here we use the sample mean as a control variable for the control-variate trimean estimator, since the sample mean is highly corelated with the trimean.

The results in the table demonstrate that the control variate trimean estimator has a significant reduction on the variance of the trimean estimator. In particular, there is about 85% reduction for the standard normal, about 81% and 62% reduction for the t-distribution with 4 and 10 degrees of freedom, repectively. The amount of reduction is similar between the standard normal with the t-distribution with 10 degrees of freedom. This is because the standard normal is more similar with t-distribution with 10 degrees of freedom than t-distribution with 4 degrees of freedom.
