---
title: "pareto-intro"
date: "2024-05-14"
output:
  rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{pareto-intro}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


The Pareto distribution density is calculated using the following formula: $$ f(x|\alpha,\beta) = \frac{\beta{\alpha}^\beta}{x^{\beta+1}} $$
for $0<\alpha\leq x$ and $\beta>0$. Note that densities are defined on the entire real line; they are zero outside of the support of the distribution.

```{r}
knitr::opts_chunk$set(fig.width = 5, fig.height = 4)
```

## Code
```{r, eval=FALSE}
Rcpp::sourceCpp("../Rcpp/dpareto.cpp")
source("../R/dpareto.R")

ppareto <- function(q, alpha, beta, lower.tail = TRUE, log.p = FALSE){
  max_length <- max(length(q), length(alpha), length(beta))
  alpha <- rep(alpha, length = max_length)
  beta <- rep(beta, length = max_length)
  q <- rep(q, length = max_length)
  log_p <- suppressWarnings(ifelse(q<alpha, 0, beta*(log(alpha)-log(q))))
  log_p <- suppressWarnings(ifelse(alpha <=0 | beta <=0, NaN, log_p))
  if (any( is.nan( log_p )) == TRUE){
    warning("NaNs produced")
  }
  p <- log_p
  if(! log.p) p <- exp(log_p)
  if(lower.tail) p <- 1 - p
  if(log.p & (! lower.tail)) p <- log(1-exp(log_p))
  p
}

qpareto <- function(p, alpha, beta, lower.tail = TRUE, log.p = FALSE){
  max_length <- max(length(p), length(alpha), length(beta))
  alpha <- rep(alpha, length = max_length)
  beta <- rep(beta, length = max_length)
  p <- rep(p, length = max_length)
  if(log.p) p = exp(p)
  if(lower.tail) p = 1 - p
  q <- suppressWarnings(alpha/(p^(1/beta)))
  q <- suppressWarnings(ifelse(alpha <=0 | beta <=0, NaN, q))
  if (any( is.nan( q )) == TRUE){
    warning("NaNs produced")
  }
  q
}

rpareto <- function(n, alpha, beta){
  qpareto(runif(n), alpha, beta)
}

```

```{r}
## R CMD INSTALL pareto_version.tar.gz
library(pareto)
```

## Usage

`dpareto` gives the density,  
`ppareto` gives the distribution function,  
`qpareto` gives the quantile function,  
`rpareto` generates random deviates. 

```{r}
dpareto(x = 2:4, alpha = 1, beta = 2)
ppareto(q = 2:4, alpha = 1, beta = 2)
qpareto(p = c(0.1, 0.5, 0.9), alpha = 1, beta = 2)
rpareto(n = 10, alpha = 1, beta = 2)
```


\newpage

## Density
```{r}
suppressMessages(library(tidyverse))
df_p <- data.frame(q = seq(1, 10, 0.1), alpha = 1, beta = 1)
df_p <- df_p %>% mutate(beta = 2) %>% rbind(df_p, .)
df_p <- df_p %>% mutate(alpha = 2) %>% rbind(df_p, .)
df_p <- df_p %>% mutate(den = dpareto(q, alpha, beta), cum = ppareto(q, alpha, beta)) %>% mutate(alpha = as.factor(alpha), beta = as.factor(beta))
```

```{r}
ggplot(df_p, aes(x = q, y = den, color = alpha, linetype = beta)) + 
  geom_line() + 
  theme_bw()
```

\newpage

## Cumulative Distribution
```{r}
ggplot(df_p, aes(x = q, y = cum, color = alpha, linetype = beta)) + 
  geom_line() + 
  theme_bw()
```


```{r eval=FALSE, include=FALSE}
## for creating R package
working.Rdir <- file.path("packages", "working", "R")
working.Rfiles <- file.path(working.Rdir, c("dpareto.R", "ppareto.R", "qpareto.R", "rpareto.R"))
package.path <- file.path("packages")
# package.skeleton("pareto", code_files = working.Rfiles, path = package.path)
Rcpp::compileAttributes(pkgdir = "packages/pareto/",verbose = TRUE)
# Rcpp::compileAttributes
## R CMD build pareto
## R CMD check pareto_1.0.tar.gz
## R CMD INSTALL pareto_1.0.tar.gz
```

\newpage

# Acknowledgement

This document depended on the packages

* Rcpp [@Rcpp-article, @Rcpp-book, @Rcpp-brief]
* markdown [@R-markdown]
* knitr [@R-knitr]

and the R environment [@R-man] to get it done.


# References
